Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-03-10 supplement.tex
edits to captions up to S22

\subsubsection{Conditional SFS}

The single-origin model fit the cSFS relatively poorly across the three
populations compared to the models that allow structure between stem
populations (Fig.~\ref{fig:supp-csfs-single-origin}). The model that best fit
the cSFS was the merger-with-stem-migration model, which also provided the best
fit to pairwise diversity and LD statistics in the initial optimization. For
the Mende and Gumuz, the merger-with-stem-migration model fits the data very
well, while it underestimates rare variants in the Nama cSFS
(Fig.~\ref{fig:supp-csfs-merger-with-stem-migration}). Models learned from the
two-locus statistics provided a better qualitative description of the cSFS than
models previously fit directly to the cSFS  (see, e.g., figure 1 in
\citet{Durvasula2020-td}). We found that the inferred ancestral
misidentification rates were concordant across populations. The inferred rates
of ancestral misidentification were about twice as high for transitions
relative to transversions as expected, qualitatively, from the higher
transition mutation rate \citep{Hernandez2007-mf}.

@@

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-csfs-single-origin.pdf}
    \caption{
        \textbf{Conditional SFS compared to single-origin model.}
        The single-origin model provides a poor fit to the observed cSFS in
        the Mende, Gumuz, and Nama. Ancestral state misidentification
        (labeled as $p_m$ in the figure legends) was
        inferred to be between 1 and $2\%$, and misidentification rates were
        consistent across model comparisons. Misidentification rates of
        transition-type mutations were roughly double that of transversion-type
        mutations, consistent with the higher mutation rates of transition
        mutations vs. transversions.
    }
    \label{fig:supp-csfs-single-origin}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-csfs-continuous-migration.pdf}
    \caption{
        \textbf{Conditional SFS compared to continuous-migration model.}
        The continuous-migration model provides a better fit to the data than
        either the single-origin model or the merger-without-stem-migration
        model, although systematic deviations are still noticeable in each
        population. Ancestral state misidentification ($p_m$), as estimated
        with all models, is roughly doubled for transition-type mutations
        compared to transversions.
    }
    \label{fig:supp-csfs-continuous-migration}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-csfs-merger-without-stem-migration.pdf}
    \caption{
        \textbf{Conditional SFS compared to merger-without-stem-migration model.}
        The excess of intermediate-frequency variants in the Nama and Mende
        cSFS results in a poorer fit to the cSFS compared to the continuous-migration
        and merger-with-stem-migratino models. Along with the poorer likelihood this
        model provides to the two-locus statistics, validation usin the cSFS
        disfavors a model with strict isolation between stem populations.
        Ancestral state misidentification ($p_m$), as estimated
        with all models, is roughly doubled for transition-type mutations
        compared to transversions.
    }
    \label{fig:supp-csfs-merger-without-stem-migration}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-csfs-merger-with-stem-migration.pdf}
    \caption{
        \textbf{Conditional SFS compared to merger-with-stem-migration model.}
        The merger-with-stem-migration model, in addition to providing the best
        fit to heterozygosity and LD statistics, provided the best fit to the cSFS
        in the Nama, Gumuz, and Mende populations. Rates of ancestral state
        misidentification were consistent across each comparison, with the
        ancestral state of transversion mutations estimated to be $0.7-1.0\%$, and
        that of transition mutations estimated to be $1.9-2.0\%$. These
        misidentification rates are consistent with the known difference in mutation
        rates between transitions and transversions, as well as other SFS-based
        inferences in humans using the Thousand Genomes data and the 6-primate
        alignment (see Methods). Ancestral state misidentification ($p_m$), as estimated
        with all models, is roughly doubled for transition-type mutations
        compared to transversions.
    }
    \label{fig:supp-csfs-merger-with-stem-migration}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-relate-iicr-data-full-and-subset}
    \caption{
        \textbf{Inverse instantaneous coalescence rates inferred from the data,
        subset to focal populations in the model inferences.}
        Fig.~\ref{fig:supp-iicr-data} shows IICR curves from reconstructed
        genealogies inferred using a larger dataset (including nearly 1,200
        individuals) than the subset of populations we primarily focused on
        here. To assess the robustness of Relate-inferred genealogies and for a
        direct comparison to our inferred demographic models, we ran Relate on
        the 290 genomes from the Nama, Gumuz, Mende, Amhara/Oromo, and British.
        While the reconstructed IICR curves are qualitatively similar to those
        from the full dataset for these populations in the distant past, there
        are noticeable discrepancies over the recent history in the last 10s of
        thousands of years. Left: Relate run on full dataset, showing focal
        populations. Right: Relate run directly on the subset dataset of focal
        populations.
    }
    \label{fig:supp-iicr-data-subset}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-relate-iicr-simulation}
    \caption{
        \textbf{Inverse instantaneous coalescence rates reinferred from
        simulated data.} We simulated data under our four highlighted models,
        sampling the same number of individuals per population as the original
        dataset, and then ran Relate on each simulated dataset to reconstruct
        IICR curves for each. Each model has qualitatively different histories,
        particularly for the distant past with varying numbers and timings of
        oscillations. However, due to the discrepancies in reconstructed
        genealogies using even the same data as input
        (Figs.~\ref{fig:supp-iicr-data-subset} and \ref{fig:supp-rccr-data}),
        it is difficult to draw firm conclusions from qualitative comparisons
        between reconstructed genealogies.
    }
    \label{fig:supp-iicr-sim}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-relate-rccr-data-full-and-subset}
    \caption{
        \textbf{Relative cross-coalescence rates inferred from the data.} Using
        the same reconstructed genealogies from Relate, we computed RCCR curves
        for pairs of populations using genealogies from the full dataset
        ($\sim1,200$ individuals) and the subset of focal populations (290
        individuals). Despite the overlap in samples, the two sets of inferred
        genealogies provided inconsistent RCCR, with both the timing and
        ordering of increased RCCR varying between the two datasets.
        Genealogies reconstructed from the full dataset provide a better match
        to each of inferred demographic models
        (Figs.~\ref{fig:supp-rccr-single-origin}--\ref{fig:supp-rccr-merger-with-stem-migration}).
    }
    \label{fig:supp-rccr-data}
\end{figure}

>>>>

\subsubsection{Conditional SFS}
\label{sec:cSFS}
The single-origin model fit the cSFS relatively poorly across the three
populations compared to the models that allow structure between stem
populations (Fig.~\ref{fig:supp-csfs-single-origin}). The model that best fit
the cSFS was the merger-with-stem-migration model, which also provided the best
fit to pairwise diversity and LD statistics in the initial optimization. For
the Mende and Gumuz, the merger-with-stem-migration model fits the data very
well, while it underestimates rare variants in the Nama cSFS
(Fig.~\ref{fig:supp-csfs-merger-with-stem-migration}). Models learned from the
two-locus statistics provided a better qualitative description of the cSFS than
models previously fit directly to the cSFS  (see, e.g., figure 1 in
\citet{Durvasula2020-td}). We found that the inferred ancestral
misidentification rates were concordant across populations. The inferred rates
of ancestral misidentification were about twice as high for transitions
relative to transversions as expected, qualitatively, from the higher
transition mutation rate \citep{Hernandez2007-mf}.

@@

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-csfs-single-origin.pdf}
    \caption{
        \textbf{Conditional SFS compared to single-origin model predictions.}
        The single-origin model provides a poor fit to the Neanderthal-conditional SFS in
        the Mende, Gumuz, and Nama (see Section \ref{sec:cSFS}). Ancestral state misidentification
        (labeled as $p_m$ in the figure legends) was
        inferred to be between 1 and $2\%$, and misidentification rates were
        consistent across model comparisons. Misidentification rates of
        transition-type mutations were roughly double that of transversion-type
        mutations, consistent with the higher mutation rates of transition
        mutations vs. transversions.
    }
    \label{fig:supp-csfs-single-origin}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-csfs-continuous-migration.pdf}
    \caption{
        \textbf{Conditional SFS compared to continuous-migration model predictions.}
        The continuous-migration model provides a better fit to the Neanderthal-conditional SFS  than
        either the single-origin model or the merger-without-stem-migration
        model, although systematic deviations are still noticeable in each
        population. Ancestral state misidentification ($p_m$), as estimated
        with all models, is roughly doubled for transition-type mutations
        compared to transversions.
    }
    \label{fig:supp-csfs-continuous-migration}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-csfs-merger-without-stem-migration.pdf}
    \caption{
        \textbf{Conditional SFS compared to merger-without-stem-migration model predictions.}
        The excess of intermediate-frequency variants in the Nama and Mende
        cSFS results in a poorer fit to the cSFS compared to the continuous-migration
        and merger-with-stem-migratino models. Thus the model without stem migrations
        provides a poorer fit 
        to both two-locus statistics and the cSFS.
        Ancestral state misidentification for transition-type mutations ($p_m$), as estimated
        with all models, is roughly double that of transversions.
    }
    \label{fig:supp-csfs-merger-without-stem-migration}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-csfs-merger-with-stem-migration.pdf}
    \caption{
        \textbf{Conditional SFS compared to merger-with-stem-migration model predictions.}
        The merger-with-stem-migration model, in addition to providing the best
        fit to heterozygosity and LD statistics, provided the best fit to the cSFS
        in the Nama, Gumuz, and Mende populations. Rates of ancestral state
        misidentification were consistent across each comparison, with the
        ancestral state of transversion mutations estimated to be $0.7-1.0\%$, and
        that of transition mutations estimated to be $1.9-2.0\%$. These
        misidentification rates are consistent with the known difference in mutation
        rates between transitions and transversions, as well as other SFS-based
        inferences in humans using the Thousand Genomes data and the 6-primate
        alignment (see Methods). 
    }
    \label{fig:supp-csfs-merger-with-stem-migration}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-relate-iicr-data-full-and-subset}
    \caption{
        \textbf{Inverse instantaneous coalescence rates inferred from the data,
        subset to focal populations in the model inferences.}
        Fig.~\ref{fig:supp-iicr-data} shows IICR curves from reconstructed
        genealogies inferred using a larger dataset (including nearly 1,200
        individuals) than the subset of populations we primarily focused on
        here. To assess the robustness of Relate-inferred genealogies and for a
        direct comparison to our inferred demographic models, we ran Relate on
        the 290 genomes from the Nama, Gumuz, Mende, Amhara/Oromo, and British.
        While the reconstructed IICR curves are qualitatively similar to those
        from the full dataset for these populations in the distant past, there
        are noticeable discrepancies over the recent history in the last tens of
        thousands of years. Left: Relate run on full dataset, showing focal
        populations. Right: Relate run directly on the subset dataset of focal
        populations. \sgcomment{should the labels on right and left panels be identical?}
    }
    \label{fig:supp-iicr-data-subset}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-relate-iicr-simulation}
    \caption{
        \textbf{Inverse instantaneous coalescence rates reinferred from
        simulated data.} We simulated data under our four highlighted models,
        sampling the same number of individuals per population as the original
        dataset, and then ran Relate on each simulated dataset to reconstruct
        IICR curves for each. Each model has qualitatively different histories,
        particularly for the distant past with varying numbers and timings of
        oscillations. However, due to the discrepancies in reconstructed
        genealogies using even the same data as input
        (Figs.~\ref{fig:supp-iicr-data-subset} and \ref{fig:supp-rccr-data}),
        it is difficult to draw firm conclusions from quantitative comparisons
        between reconstructed genealogies.
    }
    \label{fig:supp-iicr-sim}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-relate-rccr-data-full-and-subset}
    \caption{
        \textbf{Relative cross-coalescence rates inferred from the data.} Using
        the same \sgcomment{same was what? Figure S19 presumably?} reconstructed genealogies from Relate, we computed RCCR curves
        for pairs of populations using genealogies from the full dataset
        ($\sim1,200$ individuals) and the subset of focal populations (290
        individuals). Despite the overlap in samples, the two sets of inferred
        genealogies provided inconsistent RCCR, with both the timing and
        ordering of increased RCCR varying between the two datasets.
        Genealogies reconstructed from the full dataset provide a better match
        to each of inferred demographic models
        (Figs.~\ref{fig:supp-rccr-single-origin}--\ref{fig:supp-rccr-merger-with-stem-migration}).
    }
    \label{fig:supp-rccr-data}
\end{figure}
